---
title: "Introduction to decision trees using rpart package"
author: "Hicham Zmarrou"
date: "`r Sys.Date()`"
output:
  html_notebook:
    highlight: pygments
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
  word_document:
    toc: yes
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='img/',
                      echo=TRUE, warning=FALSE, message=FALSE,digits = 3)
library(tidyverse)
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)

```

## Stage C prostate cancer 

* Example is based on a data set of 146 stage C prostate cancer patients. 

* Main clinical endpoint of interest is:

   * whether the disease recurs after initial surgical removal of the prostate, and 
   * the time interval to that progression (if any). 
   
* The target variable  in this example is status, which takes on the value 1 if the disease has progressed and 0 if not.

* The main predictor variable of interest in this study was DNA ploidy, as determined by experimental studies.

* A short description of each of the variables is listed below. 

```
pgtime: Time to progression or last follow-up (years)

pgstat: 1 = progression observed, 0 = censored

age: age in years

eet: early endocrine therapy, 1 = no, 2 = yes

g2: percent of cells in G2 phase, as found by flow cytometry

grade: grade of the tumor, Farrow system

gleason: grade of the tumor, Gleason system

ploidy: the ploidy status of the tumor, from flow cytometry. Values are diploid, tetraploid, and aneuploid
```

### Explore the data by plotting

1. plotting progression rate by value of ploidy;

```{r}
# count the number of  1 = progression observed, and 0 = censored in the pgstat variable
stagec %>% 
  group_by(pgstat) %>% 
  summarize(count =n()) %>% 
  mutate(freq = count / sum(count))
```


```{r}
(counts_pldy <- stagec %>% 
  group_by(pgstat,ploidy) %>% 
  summarize(count =n()) )

#(counts <- table(clean_titanic$survived, clean_titanic$sex))
#barplot(counts, xlab = "Gender", ylab = "Number of People", main = "survived and deceased per gender")
```

```{r}
ggplot(data=counts_pldy, aes(x=ploidy,y=count,fill=factor(pgstat))) +
  geom_bar(stat="identity", position=position_dodge())+
  #geom_text(aes(y=label_ypos, label=len), vjust=1.6, 
   #         color="white", size=3.5)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal()
```





2. Claculate the progression rate by value ofearly endocrine therapy;



```{r}
(counts_eet <- stagec %>% 
  group_by(pgstat,eet) %>% 
  summarize(count =n()) )

#(counts <- table(clean_titanic$survived, clean_titanic$sex))
#barplot(counts, xlab = "Gender", ylab = "Number of People", main = "survived and deceased per gender")
```

```{r}
ggplot(data=counts_eet, aes(x=eet,y=count,fill=factor(pgstat))) +
  geom_bar(stat="identity", position=position_dodge())+
  #geom_text(aes(y=label_ypos, label=len), vjust=1.6, 
   #         color="white", size=3.5)+
  scale_fill_brewer(palette="Paired")+
  theme_minimal()
```

### Fitting a decision tree

* We now fit a tree model by using the `rpart` function from the `rpart` package



* The first argument of the function is a model formula, with the `???` symbol standing for "is modeled as". 

* The `print` function gives an abbreviated output, as for other S models. The plot and text command plot the tree and then label the plot, the result is shown in figure 3.

```{r}
progstat <- factor(stagec$pgstat, levels = 0:1, labels = c("No", "Prog"))
cfit <- rpart(progstat ~ age + eet + g2 + grade + gleason + ploidy,
data = stagec, method = 'class')

#cfit <- rpart(progstat ~ age + eet + g2 + grade + gleason + ploidy, data = stagec, method = "class", 
#                parms = list(split = 'information'), minsplit = 2, minbucket = 1)
print(cfit)
```

```{r}
fancyRpartPlot(cfit)
```

```{r}
asRules(cfit)
```

```{r}
printcp(cfit)
```




```{r}
summary(cfit, cp=0.06)
```

