---
title: "Model Basics"
author: "Exercises"
date: "`r Sys.Date()`"
output:
  html_notebook:
    highlight: pygments
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
  word_document:
    toc: yes
---


### Exercises



1. Import the advertising dataset from this [link](http://www-bcf.usc.edu/~gareth/ISL/Advertising.csv')
```{r}
Advertising<-  read_csv('http://www-bcf.usc.edu/~gareth/ISL/Advertising.csv')
```

2. Produce a scatter plot taht show the relation ship between (x = TV, y = sales)  

```{r}
ggplot(data = Advertising) + geom_point(mapping = aes(x = TV, y = sales),color = "blue")
```

2. Find the best linear model (use `lm` method/function)  

```{r}
mod1 <- lm(sales ~ TV, data= Advertising)
summary(mod1)
```


3. Make an evenly grid of the `TV` values

```{r}
grid <- Advertising %>% data_grid(TV) 
grid
```


4. Predict the `sales` values on the grid using the model
```{r}
grid <- grid %>%  add_predictions(mod1) 
grid
```


5. Visualize the linear model 
```{r}
ggplot(Advertising, aes(x=TV,y=sales)) + geom_point(colour = "blue") +
  geom_line(aes(y = pred), data = grid, colour = "red", size = 1)
```


6. Show the residuals in a table/dataframe 
```{r}
Advertising<-Advertising %>% add_residuals(mod1)

```


7. Draw the distribution of the residuals
```{r}
binwd = (max(Advertising$resid) - min(Advertising$resid))/20
ggplot(Advertising, aes(resid)) + 
  geom_freqpoly(binwidth = binwd)

```


8. Draw at the patterns of the residuals against the TV values 

```{r}
ggplot(Advertising, aes(TV, resid)) + 
  geom_ref_line(h = 0) +
  geom_point()

```

