---
title: "Model Basics"
author: "Exercises I"
date: "`r Sys.Date()`"
output:
  html_notebook:
    highlight: pygments
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
  word_document:
    toc: yes
---

### Exercises 

```{r}
suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(purrr)))
suppressWarnings(suppressMessages(library(modelr)))
```
1Q.  One downside of the linear model is that it is sensitive to unusual values
    because the distance incorporates a squared term. Fit a linear model to 
    the simulated data below, and visualise the results. Rerun a few times to
    generate different simulated datasets. What do you notice about the model? 
    
```{r echo=TRUE}
    sim1a <- tibble(
      x = rep(1:10, each = 3),
      y = x * 1.5 + 6 + rt(length(x), df = 2) 
    )

```

1.A 
Run linear model and visualize it

```{r echo=TRUE}

fit <- lm(y~x, data = sim1a)
ggplot(sim1a,aes(x,y))+
  geom_point(size = 2, color = "grey30")+
  geom_abline(intercept = fit$coefficients[1],slope = fit$coefficients[2])
```

Sometimes, one single abnormal value forces the fitted line deviate from the "intutively" best lines.

2.  One way to make linear models more robust is to use a different distance
    measure. For example, instead of root-mean-squared distance, you could use
    mean-absolute distance:
    
```{r echo=TRUE}
model1 <- function(a, data) {
  a[1] + data$x * a[2]
}

measure_distance <- function(mod, data) {
  diff <- data$y - model1(mod, data)
  mean(abs(diff))
}

measure_distance2 <- function(mod, data) {
  diff <- data$y - model1(mod, data)
  sqrt(mean(diff ^ 2))
}
```

Compare the two measures of distance
```{r echo=TRUE}
sim1a <- tibble(
  x = rep(1:10, each = 3),
  y = x * 1.5 + 6 + rt(length(x), df = 2)
)
best1 <- optim(c(0,0), measure_distance, data = sim1a)
best2 <- optim(c(0,0), measure_distance2, data = sim1a)

ggplot(sim1a, aes(x, y)) + 
  geom_point(size = 2, colour = "grey30") + 
  geom_abline(intercept = best1$par[1], slope = best1$par[2], color = "red")+
  geom_abline(intercept = best2$par[1],slope = best2$par[2], color = "cyan")
```
When there are many abnormal points, the `cyan` line will perform better using absolute distances. It is better because measn-square-distance tends to overemphasize abnormal values.


