---
title: "Data Structures Exercises"
author: "Hicham Zmarrou"
date: "Notebook -- <http://bit.ly/2pHMYck>  <br /> <br />"
output:
  html_notebook:
    highlight: pygments
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
subtitle: Introduction to R
venue: TNO data science courses
---


<hr>


__References__

Most of this material is borrowed from:

* Textbook: [R for Data Science: Import, Tidy, Transform, Visualize, and Model Data ](http://amzn.to/2r3x3ZF)

______________________________________________________________________________________________________________________________________


## Data Structures Exercises
**Exercise 1:** 
Create an atomic vector of Character type, double type, logical type , integer type, complex type and raw type


 
**Exercise 2:** 
Check whether each of the created data structure is of vector type as well as check the class of each of the data.


**Exercise 3:** 
Create a List of heterogeneous data, which include numeric, character and logical vectors and prints the lists.

**Exercise 4:** 
Create a matrix of 3 rows and 4 columns which stores data from 1 to 12 and arrange the value row wise.

**Exercise 5:** 
Create a matrix of 3 rows and 4 columns which stores random data and arrange the matrix column wise

**Exercise 6:** 
Create two 2 x 2 matrices with random values and perform all arithmetic operations with those two and print the resultant matrices

**Exercise 7:** 
Create a random matrix having values between 1 and 1000 and print the matrix and transpose of the matrix

**Exercise 8:**  Create Data Frames which contain details of 5 employees and display the same.

**Exercise 9:** 
Create Data Frames which contain details of 5 employees and display summary of the data
 
**Exercise 10:** 
Create a Data Frame of 5 employees and display the details of First and Last Employee and Names of All employees.



## Data structures Solutions 





####################
#                  #
#    Exercise 1    #
#                  #
####################

```{r}
charactertype<-'A'
integertype<-10
logicaltype<-TRUE
complextype<-2+3i
rawtype<-charToRaw('x')

###

print(charactertype)
print(integertype)
print(logicaltype)
print(complextype)
print(rawtype)

```

####################
#                  #
#    Exercise 2    #
#                  #
####################

```{r}
is.vector(charactertype)

is.vector(integertype)

is.vector(logicaltype)

is.vector(complextype)

is.vector(rawtype)

class(charactertype)

class(integertype)

class(logicaltype)

class(complextype)

class(rawtype)
```



####################
#                  #
#    Exercise 3    #
#                  #
####################
```{r}
list1<-list(No = 1:10,Chr="Hello",Flag=TRUE)
print(list1)
```


####################
#                  #
#    Exercise 4    #
#                  #
####################
```{r}
Mat <- matrix(3:14, nrow = 4,ncol=3,byrow = TRUE)
print(Mat)
```

####################
#                  #
#    Exercise 5    #
#                  #
####################

```{r}
Mat <- matrix(c(2,3,6,7,8,9,12,22,-5,98,56,-77), nrow = 3,ncol=4)
print(Mat)
```
####################
#                  #
#    Exercise 6    #
#                  #
####################

```{r}
matrix1<-matrix(sample(1:10,4),nrow=2,ncol=2,byrow=TRUE)
print("Matrix 1 with random value between 1 and 10")

print(matrix1)

matrix2<-matrix(sample(1:10,4),nrow=2,ncol=2,byrow=TRUE)
print("Matrix 2 with random value between 1 and 10")

print(matrix2)

summatrix<-matrix1+matrix2
submatrix<-matrix1-matrix2
multmatrix<-matrix1*matrix2
divmatrix<-matrix1/matrix2
print("sum of the Matrices")

print(summatrix)

print("Difference of the Matrices")

print(submatrix)

print("Multiplaction of the Matrices")

print(multmatrix)

print("Division of the Matrices")

print(divmatrix)

```

####################
#                  #
#    Exercise 7    #
#                  #
####################

```{r}
matrix1<-matrix(sample(1:1000,9),nrow=3,ncol=3,byrow=TRUE)
print("Matrix 1 with random value between 1 and 1000")

print(matrix1)

print("Transposed Matrix")

print(t(matrix1))
```

####################
#                  #
#    Exercise 8    #
#                  #
####################
```{r}
Employees<-data.frame(Name=c("ALAN S","RYAN S","SERAH S", "CHRISTY S","THOMAS MARTIN"),
                      Gender=c("Male","Male","Female","Female","Male"),
                      Age=c(23,22,25,26,32),
                      Designation=c("Clerk","Manager","Exective","CEO","CTO"),
                      SSN=c("134-34-2345","349-44-789","556-34-443","898-98-987","679-67-676")
                      )
print(Employees)
```


####################
#                  #
#    Exercise 9    #
#                  #
####################

```{r}
Employees<-data.frame(Name=c("ALAN S","RYAN S","SERAH S", "CHRISTY S","THOMAS MARTIN"),
                      Gender=c("Male","Male","Female","Female","Male"),
                      Age=c(23,22,25,26,32),
                      Designation=c("Clerk","Manager","Exective","CEO","CTO"),
                      SSN=c("134-34-2345","349-44-789","556-34-443","898-98-987","679-67-676")
)
print(summary(Employees))
```


####################
#                  #
#    Exercise 10   #
#                  #
####################

```{r}
Employees<-data.frame(Name=c("ALAN S","RYAN S","SERAH S", "CHRISTY S","THOMAS MARTIN"),
                      Gender=c("Male","Male","Female","Female","Male"),
                      Age=c(23,22,25,26,32),
                      Designation=c("Clerk","Manager","Exective","CEO","CTO"),
                      SSN=c("134-34-2345","349-44-789","556-34-443","898-98-987","679-67-676")
)
print("Employee 1 Details")

print(Employees[1,])

print("Employee 5 Details")

print(Employees[5,])

print("Name of all employees")

print(Employees[1])
```