---
title: "SQL databases and R"
author: "Hicham Zmarrou"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
  html_notebook:
    highlight: pygments
    number_sections: no
    theme: cosmo
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
---


## Introduction to dbplyr

As well as working with local in-memory data stored in data frames, dplyr also works with remote on-disk data stored in databases. This is particularly useful in two scenarios:

* Your data is already in a database.

* You have so much data that it does not all fit into memory simultaneously and you need to use some external storage engine.

To use databases with dplyr you need to first install dbplyr:


```{r }

# install.packages("dbplyr")

```

You'll also need to install a `DBI` backend package. The `DBI` package provides a common interface that allows dplyr to work with many different databases using the same code. `DBI` is automatically installed with dbplyr, but you need to install a specific backend for the database that you want to connect to.

Five commonly used backends are:

* [RMySQL](https://github.com/rstats-db/RMySQL#readme) connects to MySQL and MariaDB

* [RPostgreSQL](https://cran.r-project.org/package=RPostgreSQL) connects to Postgres and Redshift.

* [RSQLite](https://github.com/rstats-db/RSQLite) embeds a SQLite database.

* [odbc](https://github.com/rstats-db/odbc#odbc) connects to many commercial databases via the open database connectivity protocol.

* [bigrquery](https://github.com/rstats-db/bigrquery) connects to Google's BigQuery.

If the database you need to connect to is not listed here, you'll need to do some investigation (i.e. googling) yourself.

In this tutorial, we're going to use the RSQLite backend which is automatically installed when you install dbplyr. SQLite is a great way to get started with databases because it's completely embedded inside an `R` package. Unlike most other systems, you don't need to setup a separate database server. SQLite is great for demos, but is surprisingly powerful, and with a little practice you can use it to easily work with many gigabytes of data.



Connecting to the database

To work with a database in dplyr, you must first connect to it, using `DBI::dbConnect()`. We're not going to go into the details of the `DBI` package here, but it's the foundation upon which dbplyr is built. You'll need to learn more about if you need to do things to the database that are beyond the scope of dplyr.

```{r}
library(tidyverse)
library(dplyr)
con <- DBI::dbConnect(RSQLite::SQLite(), path = ":memory:")
```

Another option is to connect to an sqlite database already existing on your hard disk. You may for example download and connect to the SQLite database contained in the file `portal_mammals.sqlite` in the following link:

```{r eval=FALSE, include=FALSE}
dir.create("data", showWarnings = FALSE)
download.file(url = "https://ndownloader.figshare.com/files/2292171",
              destfile = "data/portal_mammals.sqlite", mode = "wb")  
```
And point `R` to this database using:

```{r}
mammals <- DBI::dbConnect(RSQLite::SQLite(), "data/portal_mammals.sqlite",synchronous = NULL)
```


The arguments to `DBI::dbConnect` vary from database to database, but the first argument is always the database backend. It's `RSQLite::SQLite()` for `RSQLite`, `RMySQL::MySQL()` for `RMySQL`, `RPostgreSQL::PostgreSQL()` for `RPostgreSQL`, `odbc::odbc()` for `odbc`, and `bigrquery::bigquery()` for `BigQuery`. `SQLite` only needs one other argument: the path to the database. Here we use the special string `":memory:"` which causes SQLite to make a temporary in-memory database.

Most existing databases don't live in a file, but instead live on another server. That means in real-life that your code will look more like this:

```{r eval=FALSE, include=FALSE}
con <- DBI::dbConnect(RMySQL::MySQL(), 
  host = "host address",
  user = "username",
  password = ("Database password")
)
```


Our temporary database has no data in it, so we'll start by copying over nycflights13::flights using the convenient copy_to() function. This is a quick and dirty way of getting data into a database and is useful primarily for demos and other small jobs.

```{r}
copy_to(con, nycflights13::flights, "flights",
  temporary = FALSE, 
  indexes = list(
    c("year", "month", "day"), 
    "carrier", 
    "tailnum",
    "dest"
  )
)
```

As you can see, the copy_to() operation has an additional argument that allows you to supply indexes for the table. Here we set up indexes that will allow us to quickly process the data by day, carrier, plane, and destination. Creating the write indices is key to good database performance, but is unfortunately beyond the scope of this article.

Now that we've copied the data, we can use tbl() to take a reference to it:

```{r}
flights_db <- tbl(con, "flights")
```

When you print it out, you'll notice that it mostly looks like a regular tibble:

```{r}
flights_db 
```

The main difference is that you can see that it's a remote source in a SQLite database.

## Generating queries

To interact with a database you usually use SQL, the Structured Query Language. `SQL` is over 40 years old, and is used by pretty much every database in existence. The goal of dbplyr is to automatically generate `SQL` for you so that you're not forced to use it. However, `SQL` is a very large language and dbplyr doesn't do everything. It focusses on `SELECT` statements, the `SQL`you write most often as an analyst.

Most of the time you don't need to know anything about SQL, and you can continue to use the dplyr verbs that you're already familiar with:

```{r}
flights_db %>% select(year:day, dep_delay, arr_delay)

```


```{r}
flights_db %>% filter(dep_delay > 240)
```

```{r}
flights_db %>% 
  group_by(dest) %>%
  summarise(delay = mean(dep_time))
```

However, in the long-run, I highly recommend you at least learn the basics of SQL. It's a valuable skill for any data analyst, and it will help you debug problems if you run into problems with dplyr's automatic translation. If you're completely new to `SQL` you might start with this [codeacademy tutorial](https://www.codecademy.com/learn/learn-sql). If you have some familiarity with `SQL` and you'd like to learn more, I found [how indexes work in SQLite](http://www.sqlite.org/queryplanner.html) and [10 easy steps to a complete understanding of SQL](https://blog.jooq.org/2016/03/17/10-easy-steps-to-a-complete-understanding-of-sql/) to be particularly helpful.

The most important difference between ordinary data frames and remote database queries is that your `R` code is translated into `SQL` and executed in the database, not in R. When working with databases, `dplyr` tries to be as lazy as possible:

* It never pulls data into `R` unless you explicitly ask for it.

* It delays doing any work until the last possible moment: it collects together everything you want to do and then sends it to the database in one step.

For example, take the following code:
```{r}
tailnum_delay_db <- flights_db %>% 
  group_by(tailnum) %>%
  summarise(
    delay = mean(arr_delay),
    n = n()
  ) %>% 
  arrange(desc(delay)) %>%
  filter(n > 100)

```

Suprisingly, this sequence of operations never touches the database. It's not until you ask for the data (e.g. by printing `tailnum_delay`) that dplyr generates the `SQL` and requests the results from the database. Even then it tries to do as little work as possible and only pulls down a few rows.


```{r}
tailnum_delay_db
```

Behind the scenes, dplyr is translating your `R` code into SQL. You can see the `SQL` it's generating with show_query():


```{r}
tailnum_delay_db %>% show_query()
```



If you're familiar with SQL, this probably isn't exactly what you'd write by hand, but it does the job. You can learn more about the `SQL` translation in vignette("sql-translation").

Typically, you'll iterate a few times before you figure out what data you need from the database. Once you've figured it out, use collect() to pull all the data down into a local tibble:

```{r}
tailnum_delay <- tailnum_delay_db %>% collect()
tailnum_delay
```


* `collect()` requires that database does some work, so it may take a long time to complete. Otherwise, `dplyr` tries to prevent you from accidentally performing expensive query operations:

* Because there's generally no way to determine how many rows a query will return unless you actually run it, `nrow()` is always NA.

* Because you can't find the last few rows without executing the whole query, you can't use `tail()`.

```{r eval=FALSE, include=FALSE}
nrow(tailnum_delay_db)
#> [1] NA

tail(tailnum_delay_db)
#> Error: tail() is not supported by sql sources
```

You can also ask the database how it plans to execute the query with `explain()`. The output is database dependent.

## Creating your own database

If you don't already have a database, here's some advice from my experiences setting up and running all of them. `SQLite` is by far the easiest to get started with. `PostgreSQL` is not too much harder to use and has a wide range of built-in functions. In my opinion, you shouldn't bother with `MySQL/MariaDB:` it's a pain to set up, the documentation is subpar, and it's less featureful than `Postgres`. Google `BigQuery` might be a good fit if you have very large data, or if you're willing to pay (a small amount of) money to someone who'll look after your database.

All of these databases follow a client-server model - a computer that connects to the database and the computer that is running the database (the two may be one and the same but usually isn't). Getting one of these databases up and running is beyond the scope of this article, but there are plenty of tutorials available on the web.


### MySQL/MariaDB

In terms of functionality, MySQL lies somewhere between SQLite and PostgreSQL. It provides a wider range of [built-in functions](http://dev.mysql.com/doc/refman/5.0/en/functions.html), but it does not support window functions (so you can't do grouped mutates and filters).

### PostgreSQL

PostgreSQL is a considerably more powerful database than SQLite. It has:

* a much wider range of [built-in functions](http://www.postgresql.org/docs/9.3/static/functions.html), and

* support for [window functions](http://www.postgresql.org/docs/9.3/static/tutorial-window.html), which allow grouped subset and mutates to work.

### BigQuery

BigQuery is a hosted database server provided by Google. To connect, you need to provide your project, dataset and optionally a project for billing (if billing for project isn't enabled).

It provides a similar set of functions to Postgres and is designed specifically for analytic workflows. Because it's a hosted solution, there's no setup involved, but if you have a lot of data, getting it to Google can be an ordeal (especially because upload support from `R` is not great currently). 